\mytitle

### 2023-05-12 paper-seb-main.tex
Revisions for Computer Journal

@@

\noindent
Scientists rarely assure the structure and quality of the code they rely on, and they rarely make it accessible for scrutiny. Even if available, scientists rarely provide adequate documentation to understand or use their code reliably. When code used in science avoids adequate scrutiny published science will be unreliable, because results --- data, graphs, images, inferences, etc --- generated by the code will themselves be unreliable.

\hskip 1em This paper justifies and proposes ways to improve research and science relying on code:

@@

\indented 4
The Reproducible Analytic Pipeline (RAP) methodology can be generalized to cover code and Software Engineering methodologies, in a generalization this paper introduces called \RAPstarp. \RAPstar\ (or comparable interventions) could be supported and encouraged in journal, conference, and funding body policies.

\vskip 1ex 
\justifying
\noindent
The key point is that as code is now an inextricable part of science it should be supported by competent Software Engineering input, analogously to statistics being properly supported by critical and competent statistics input. The \supplement\ provides a summary of Software Engineering best practice relevant to scientific research, including further suggestions for \RAPstar\ processes.

@@

\section{Introduction}
Unreliable, often unstated and unexplored, code and computational dependencies (including using AI or ML) in science are widespread. Furthermore, code is rarely published or made accessible in a useful form, nor is it professionally scrutinized. Too often, therefore, code and results claimed do not contribute to reliable science, and do not support replication or reproduction on which future science can be firmly based.

@@

This paper makes an explicit analogy between the use of statistics, which has clear standards for reliable use and presentation, and the use of code. Like statistics, code is often relied on to support key results in scientific publications, yet code is generally informal, inaccessible and incorrect. Just as sound experimental methods and sound statistics generally rely on professional specialist input, it is argued here that good use of code must rely on professional Software Engineering input \cite{sommerville,knight}, and more strategically on Computational Thinking \cite{computational-thinking-wing,computational-thinking} --- an accessible form of Software Engineering that consciously applies the ideas universally, far more widely than just to software and coding.

\begin{table*}
\begin{center}\sf\normalsize
\begin{tabular}{|rl|} \hline
\the\numberOfJournals&Journals\\
\the\dataN&Papers:\\
\tabularJournalBreakdown
\the\countAuthors&Published authors\\
\the\totalPages&Published journal pages\\
July 2020&Sample month\\ \hline
\end{tabular}
\end{center}
\caption{Overview of peer-reviewed paper sample, broken down further in Table \ref{table-summary}. Survey methodology and data is provided in the \supplement. (The survey does not include the motivating papers \cite{ICmodel,avianFluModel,originalICmodel}, none of which provide code or code summaries; see section \ref{section-pandemic-modeling}.)}
\label{table-overview}
\end{table*}

\begin{table*}
\begin{center}\sf\normalsize
\input generated/summary-table.tex
\end{center}
\caption{Breakdown of pilot survey of peer-reviewed science papers relying on code. }
\label{table-summary}
\end{table*}

\begin{table*}
\newdimen\quarterwidth
\newdimen\separation
\separation=3mm
\quarterwidth=\textwidth
\advance\quarterwidth by -4\separation
\divide\quarterwidth by 4
\noindent{\sf\begin{tabular}{@{}c|c|c|c@{}}
\bfseries Level 0&
\bfseries Level 1&
\bfseries Level 2&
\bfseries Level 3 \\ \hline
\multicolumn{1}{@{}p{\quarterwidth}}{Journal encourages code sharing -- or says nothing.}&
\multicolumn{1}{|p{\quarterwidth}}{Article states whether code is available and, if so, where to access them.}&
\multicolumn{1}{|p{\quarterwidth}}{Code must be posted to a trusted repository. Exceptions must be identified at article submission.}&
\multicolumn{1}{|p{\quarterwidth}@{}}{Code must be posted to a trusted repository, and reported analyses will be reproduced independently before publication.}
\end{tabular}}
\caption{The TOP committee's recommended levels for journal article code transparency. Level 0 is provided for a comparison that does not meet any TOP requirements. Concerns about the interpretation of ``reproduced independently,'' as required at level 4, are raised in section \ref{reproducibility-concerns}.}
\label{TOPtable}
\end{table*}

This paper was initially motivated by concerns about the poor quality of code used in high-profile epidemiology research, because of its significance for informing and driving public health responses to the Covid pandemic. A pilot survey implies that such problems are ubiquitous and by no means limited to epidemiology: see the tables \ref{table-overview} \& \ref{table-summary} for a summary of the sample, and see the \supplement\ for further details of the survey. In the survey, \emph{no\/} papers claimed or provided evidence that their code was adequately tested or rigorously developed; \emph{none\/} used methodologies like RAP or \RAPstar\ (described below). Only one paper mentioned any Software Engineering methods, albeit without technical details.

In the survey sample 81\% of papers were published in journals that have code policies (which themselves are weak), but 42\% of surveyed papers published in those journals breached their own policies. One paper declared it had accessible code, but the relevant repository was and still remains empty. The findings are comparable to problems increasingly recognized for data and data access (reviewed in section \ref{related-work}): code problems form part of the \emph{reproducibility crisis\/} \citeeg{reproducibility-crisis,rescience}.

This paper therefore argues that code should be developed and discussed in a professional, rigorous, and supportive environment that facilitates quality science with clear presentation and appropriately rigorous scrutiny of code. Its main contribution is to suggest straightforward ways to enable this. The proposals may not be ``the'' right or best ways, but it is hoped the case studies and arguments presented here persuade readers that the proposals are at least a productive way to start pointing in the right direction, and inform raising the profile and constructively debating the issues more widely.

@@

\section{Background}
The discoveries and inventions of scientific technologies and instruments like microscopes, telescopes, and X-rays, drive and expand the sciences. There are fascinating periods as new ideas and science unify; for example, thermometers could not measure temperature in any meaningful way until the science was mature. For over a century, there was no agreement on definitions of temperature, how to calibrate thermometers, or what units they measured in.

Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having a consensus in turn depended on reproducible thermometer measurements and a thorough understanding of the science, including all the confounding factors that were being misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate.} 
Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theoretical-and-technical maturation, until today, when we have robust off-the-shelf instruments that measure temperature in reliable, repeatable, internationally standardized units, following international quality standards.

Computers are a unique, new technology, far more flexible and challenging than thermometers. Understanding computers and integrating them into science is likely to be harder than the development of modern thermometry. Computation not only expands science's paradigms and supports new discoveries (particularly with AI), but it also \emph{does\/} new science --- almost all modern laboratory instrumentation, including thermometers, is heavily computerized. Pushing the boundaries of science, then, involves pushing the boundaries of computer science. The synergy runs deep: for instance, while particle physics relies on powerful supercomputers, quantum physics itself is developing more powerful quantum computing.

``Computational science'' has come to mean a particular style of science based on developing and using explicit computational models, but, really, \emph{all\/} of science is now computational in this sense. Computational science is not just restricted to specialized fields like computational chemistry, genomics, big data~\ldots\ In all fields of science, computation is used at every step, from calculations of course, through note taking, sound and image processing, literature searches, analysis and statistics, correspondence with co-authors and editors, through to typesetting,  distributing and archiving the final publications. All areas of science are being profoundly computerized. Furthermore, developments in computer science themselves drive science such that earlier science is even becoming obsolete as the computer technology moves on \cite{form}.

\subsection{Related code quality concerns}\label{related-work}
Publishing high quality computer code has been strongly advocated since the earliest times, such as the \emph{Communications of the ACM\/} in its first issue in its first volume published in 1958, where it outlined its algorithm publication policy. The new policy was illustrated with a square root algorithm \cite{acm-algorithms}. However, publishing code in the computer science literature is distinct from publishing high quality general science that depends on code, which is the concern of the present paper. Of course, as a special case, Computer Science too can also benefit from improving ways to reliably use code.

In almost all published science, the code it relies on is taken for granted, just as in routine chemistry the quality of the glassware is not at issue. While chemists are trained in reliable methodologies, so taking quality glassware for granted is reasonable. Code is a newer innovation. Reliable code is a current research programme in its own right, and has resulted in calls for a Grand Challenge research effort \cite{hoare-correctness}. Inevitably, because of these reasons taken collectively, much published science depends on unreliable code that is not explicitly discussed, was not peer-reviewed, and is not open to scrutiny, reproduction, or reuse by the scientific community.

A study of 863,878 Python-coded Jupyter notebooks \cite{jupyter-study} found a 76\% failure rate for code to complete execution successfully. Trisovic \emph{et al.}\ \cite{trisovic} performed a  study of 9,000 research codes written in the language R on Dataverse, an open-source repository maintained by Harvard University's Institute for Quantitative Social Sciences. They found a comparable result that 74\% of the code files analyzed failed. These results are consistent with this paper's findings, as summarized in tables \ref{table-overview} \& \ref{table-summary}.

The authors of \cite{jupyter-study} make technical recommendations to improve reproducibility, such as ``Abstract code into functions, classes, and modules and test them.'' More generally, the authors of \cite{trisovic} recommend establishing Working Groups to support reproducible research --- and in fact Dataverse has already done so. These ideas may be compared to the present paper's proposal of Software Engineering Boards (SEBs), as discussed in detail in section \ref{summary}, supported by more specific suggestions in the \supplement.

These concerns about code are part of the reproducibility crisis for science generally \cite{machine-learning-reproducibility,no-raw-data,data-access,reproducible-manifesto,reproducibility-crisis,reproducible-manifesto}. Concern has led to a new journal \emph{ReScience~C\/} to explore and encourage the explicit replication of previously published research \cite{rescience}; furthermore, \emph{ReScience~C\/} promotes new and open-source implementations in order to ensure that the original research is reproducible. However, most of its published concern focuses on data rather than code.

To start to address code quality issues, the Transparency and Openness Promotion Committee first met in 2014, and has since been promoting Transparency and Openness Promotion, TOP, starting with journal publication policies \cite{TOP}. TOP covers citation standards, replication standards, and code standards amongst others. TOP recommends levels of compliance to their recommendations, where level 0 does not meet the standard, and levels 1 to 3 are increasingly stringent. The TOP levels for code are shown in table \ref{TOPtable}. The TOP standards  continue to develop, and are now maintained on a wiki at \url{osf.io/9f6gx/wiki/Guidelines} \cite{TOP-wiki}. Note that TOP aligns with but is a stricter approach than the Findable, Accessible, Interoperable and Reusable (FAIR) initiative, which is critiqued in section \ref{critique-fair}.

As the present paper argues, developing quality code is widely under-appreciated, which leads to a vicious cycle of lack of acknowledgement, invisibility, and being unable to recruit adequately competent coders (section \ref{section-vicious-cycles}). The term \emph{research software engineer\/} was coined in 2012 to help address this problem, and to stimulate thinking about researchers' career paths. The Society of Research Software Engineering (\url{https://society-rse.org}) has been established to further promote research software engineer interests.

There is no shortage of computational tools available to help address the problems. However, it should be noted that such tools are not a panacea, as the study \cite{jupyter-study} cited above makes clear. This suggests that human support for improving coding and reproduction quality, perhaps in the form of Working Groups or Boards, as this paper suggests, will be important.

In addition to unintentional problems with code quality and reproducibility, scientific misconduct occurs when the outcome is intentional. While pure plagiarism, which is misconduct, generally does not affect the quality of reported science, when data or code is fraudulently manipulated to have realistic properties, the results are likely to be destructive. The Wakefield MMR fraud linking vaccines and autism published in \emph{The Lancet\/} took 12 years before it was retracted, and has been extraordinarily destructive \cite{MMR-fraud}.

A recent meta-analysis of surveys of scientific misconduct estimates that nearly 2\% of scientists at least once fabricated, falsified, or modified results, and over a third undertook other questionable research practices \cite{misconduct-issues}. The meta-analysis qualifies the figures carefully, but these are alarming rates with any qualifications; indeed, the authors suggest that as misconduct is a sensitive issue, the rates are likely to be under-estimates.\footnote{The author's survey of scientists publishing in the \emph{Journal of Machine Learning\/} had comparable results \cite{machine-learning-reproducibility}.}

While technical solutions like using AI may help, it is notable that many misconduct issues can be detected and constructively managed prior to publication using the same methods as will improve research reproducibility, as discussed throughout the present paper.

\subsection{Computable papers}\label{computable-papers}
``Electronic lab notebooks'' (ELNs) \cite{elns}, emphasize computer tools that specifically support laboratory notebook authoring and editing. In contrast, ``computable papers'' aim to support the scientific paper authoring process: the emphasis is that papers should produce faithful results from code embedded in (or easily accessible from) the text of the paper.

If, for example, HTML is being used to prepare a computational paper, it could include Javascript code like \texttt{<script> document.write(responses.total) </script>}, which would insert the result of running that Javascript code into the paper, such as the number into a sentence in an empirical paper, like ``We collected data and obtained 754 responses.''

Systems like \LaTeX\ (which was used for the current paper) can combine advanced typesetting with basic calculations (e.g., here we calculate $10!=
\def\factorial#1{\ffactorial{#1}1}
\newcount\t
\def\ffactorial#1#2{%
\ifnum#1=0 #2\else
\ffactorial{\the\numexpr#1-1\relax}{\the\numexpr#1*#2\relax}%
\fi}
\factorial{10}$ by writing \LaTeX\ code \emph{in\/} this paper). More practically, a separate program can generate a file of \LaTeX\ definitions for data, tables, and cross-references, etc, as needed for a paper.

\newdimen\boxwidth
\boxwidth = \columnwidth
\advance\boxwidth by - 1em
\noindent\framebox[\columnwidth][r]{\parbox{\boxwidth}{\noindent\sf For the present paper, as a concrete example of this approach, the pilot survey analyzed \textbf{\the\dataN} papers with \textbf{\the\countAuthors} authors. The paper \textbf{\cite{ICmodel}}, discussed below in section \textbf{\ref{section-pandemic-modeling}}, relied on code composed of \textbf{\covidsimfiles} files, with over \textbf{\covidsimkLOC\ thousand} lines of code (how this compares with files in the survey is summarized in table \textbf{\ref{supplementaryTableRepoSummary}} in the \supplement).}}

In the ``old days'' such numbers (\the\dataN, \the\countAuthors, etc) quoted above would have been manually worked out, then eyeballed and typed up by the authors. Instead, in this paper, those numbers (and many more) were computed automatically. Furthermore, they update automatically when the data or calculations change. They were inserted into the text of this paper automatically, and are auditable back to their sources.

The idea is easy to implement. Code can generate text like \texttt{\textbackslash newcommand\{\textbackslash numberOfAuthors\}\{\the\countAuthors\}} (where the \the\countAuthors\ is the calculated number) save that line of text to a file, which is then input into the paper where it is typeset in the normal way. Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the correct value is at the time. Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11} \ldots\ 1,000 etc, as done (for example) in this paper's footnote~\numbernamesexamplefootnote.

Rather than use quirky languages like \LaTeX\ computable papers more often build on general-purpose programming languages. Python and R, for instance, have powerful features for processing data, plotting graphs, doing statistics, and more.

The key concept in computable papers is that as the authors of a paper collect or revise data or calculations or otherwise update it, the results in the paper update automatically, and all the statistics, graphs, and analysis reported in the paper remain correct --- \emph{with no further work from the authors}. Indeed, if an author corrects a mistake, the correction will apply automatically to all future edits.

However, there is no structure to using general-purpose systems like HTML, \LaTeX, or Rmarkdown. Many authors therefore prefer a more structured approach imposed by tools designed for the purpose. A taut review is Perkel \cite{perkel-review}, but here four tools,  WEB, \emph{Mathematica}, Jupyter, and knitr, will serve to sample the variety of approaches that are available:

@@

\hskip 1em WEB documents an entire program, but there are variants such as relit \cite{relit} that allow arbitrary parts of programs to be documented, and hence are useful for normal scientific papers that need to explain algorithms, but do not need to show or explain the entire code.

\hskip 1em In contrast to the other tools reviewed here, literate programming is intended to produce high quality publications \emph{about\/} code, rather than publications \emph{using\/} the code, inserting output, such as graphs, generated by running it.

\item
\emph{Mathematica\/} is one of the earliest fully interactive notebook tools for computational papers. Notebooks were developed by Theodore Gray in 1988 \cite{notebook-patent}. Notebooks consist of a collection of ``cells.'' Some notebook cells are labelled as text or as section headings, but if a cell is labelled as code, it can be run and it will normally generate a new cell following it as its output. The new cell can be numbers, tables, mathematics, a plot, an image, or even more text --- the arbitrary output of running code, in fact. Moreover, a notebook itself is a \emph{Mathematica\/} expression, so it --- the paper itself --- can be analyzed or manipulated by code in any way. In particular, the entire notebook structure and contents can be checked for consistency and correctness in any way the author chooses.

>>>>

\setcounter{tocdepth}{3}

@@

\noindent
Scientists rarely rigorously ensure the structure and quality of the code they rely on, and they rarely make it accessible for scrutiny. Even if available, scientists rarely provide adequate documentation to understand or use their code reliably. When code used in science avoids adequate scrutiny, published science will be unreliable and unverifiable because results --- data, graphs, images, inferences, and so forth --- informed or generated by the code will themselves be unreliable and unverifiable.

\hskip 1em This paper proposes and justifies  ways to improve research and science relying on code:

@@

\indented 4
The Reproducible Analytic Pipeline (RAP) methodology can be generalized to cover code and Software Engineering methodologies, in a generalization this paper introduces called \RAPstarp. \RAPstar\ (or comparable interventions) could be supported and or even required in journal, conference, and funding body policies.

\vskip 1ex 
\justifying
\noindent
The key point is that as code is now an inextricable part of science it should be supported by competent Software Engineering input, analogously to statistics being properly supported by critical and competent statistics input. The paper's \supplement\ provides a summary of Software Engineering best practice relevant to scientific research, including further suggestions for \RAPstar\ workflows.

@@

\section{Introduction}
Unreliable, often unstated and unexplored, code and computational dependencies (including using AI or ML) in science are widespread. Furthermore, code is rarely published or made accessible in a usable form; it is generally too onerous or impossible to verify or scrutinize. Ironically, computers should be able to make reproducibility easier, yet too often code and results claimed do not contribute to reliable science, and do not support verification, replication, or reproduction on which future science can be firmly based.

@@

This paper makes an explicit analogy between the use of statistics, which has clear standards for reliable use and presentation, and the use of code. Like statistics, code is often relied on to support key results in scientific publications, yet code is generally informal, inaccessible and incorrect. Just as sound experimental methods and sound statistics generally rely on professional specialist input, it is argued here that good use of code must rely on professional Software Engineering input \cite{sommerville,knight}, and more strategically on professional Computational Thinking \cite{computational-thinking-wing,computational-thinking} --- an accessible form of Software Engineering that consciously applies the ideas universally, far more widely than just to software and coding.

\begin{table*}
\begin{center}\sf\normalsize
\begin{tabular}{|rl|} \hline
\the\numberOfJournals&Journals\\
\the\dataN&Papers:\\
\tabularJournalBreakdown
\the\countAuthors&Published authors\\
\the\totalPages&Published journal pages\\
July 2020&Sample month\\ \hline
\end{tabular}
\end{center}\vskip -2ex
\caption{Overview of peer-reviewed paper sample, broken down further in Table \ref{table-summary}. Survey methodology and data is provided in the \supplement. (The survey does not include the motivating papers \cite{ICmodel,avianFluModel,originalICmodel}, none of which provide code or code summaries; see section \ref{section-pandemic-modeling}.)}
\label{table-overview}
\end{table*}

\begin{table*}
\begin{center}\sf\normalsize
\input generated/summary-table.tex
\end{center}\vskip -2ex
\caption{Breakdown of pilot survey of peer-reviewed science papers relying on code. }
\label{table-summary}
\end{table*}

\begin{table*}
\newdimen\quarterwidth
\newdimen\separation
\separation=4.5mm
\quarterwidth=\textwidth
\advance\quarterwidth by -4\separation
\divide\quarterwidth by 4
\begin{center}
\noindent{\sf\begin{tabular}{|@{}c|c|c|c@{}|} \hline
\bfseries Level 0&
\bfseries Level 1&
\bfseries Level 2&
\bfseries Level 3 \\ \hline
\multicolumn{1}{|p{\quarterwidth}}{Journal encourages code sharing -- or says nothing.}&
\multicolumn{1}{|p{\quarterwidth}}{Article states whether code is available and, if so, where to access them.}&
\multicolumn{1}{|p{\quarterwidth}}{Code must be posted to a trusted repository. Exceptions must be identified at article submission.}&
\multicolumn{1}{|p{\quarterwidth}|}{Code must be posted to a trusted repository, and reported analyses will be reproduced independently before publication.} \\ \hline
\end{tabular}}\end{center}\vskip -2ex
\caption{The TOP committee's recommended levels for journal article code transparency. Level 0 is provided for a comparison that does not meet any TOP requirements. Concerns about the interpretation of ``reproduced independently,'' as required at level 4, are raised in section \ref{reproducibility-concerns}.}
\label{TOPtable}
\end{table*}

This paper was initially motivated by concerns about the poor quality of code used in high-profile epidemiology research because of its significance for informing and driving public health responses to the Covid pandemic. A pilot survey implies that such problems are ubiquitous and by no means limited to epidemiology: see tables \ref{table-overview} \& \ref{table-summary} for a summary of the sample, and see the \supplement\ for further details of the survey. In the survey, \emph{no\/} papers claimed or provided evidence that their code was adequately tested or rigorously developed; \emph{none\/} used methodologies like RAP or \RAPstar\ (described below). Only one paper mentioned any Software Engineering methods, albeit simplistic and without technical details.

In the survey sample 81\% of papers were published in leading journals that have code policies (which themselves are weak), but 42\% of surveyed papers published in those journals breached their own policies. One paper declared it had accessible code, but the relevant repository was and still remains empty. The findings are comparable to problems increasingly recognized for data and data access (reviewed in section \ref{related-work}): code problems form part of the \emph{reproducibility crisis\/} \citeeg{reproducibility-crisis,rescience} discussed throughout this paper.

This paper therefore argues that code should be developed and discussed in a professional, rigorous, and supportive environment that facilitates quality science with clear presentation and appropriately rigorous scrutiny of code. Its main contribution is to suggest straightforward ways to enable this. The proposals may not be ``the'' right or best ways, but it is hoped the case studies and arguments presented here persuade readers that the proposals are at least a productive way to start pointing in the right direction, and to inform raising the profile and constructively debating the issues more widely.

@@

\section{Background}
The discoveries and inventions of scientific technologies and instruments like microscopes, telescopes, and X-rays, drove and expanded the sciences. There are fascinating periods when new ideas and science unified; for example, thermometers could not measure temperature in any meaningful way until the underlying science was mature. For over a century, there was no agreement on definitions of temperature, how to calibrate thermometers, or what units they measured in.

Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having a consensus in turn depended on reproducible thermometer measurements; for example, scientists working in different places needed to know they were working with the ``same temperatures'' yet there was for a long period no consensus on what that meant. Contributing to reliable science depended on a thorough understanding of principles, including gradually fixing the confounding factors that were misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate.} 
Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theoretical-and-technical maturation, until today, when we have robust off-the-shelf instruments that measure temperature in reliable, repeatable, internationally standardized units, that follow international quality standards.

Computers are a unique, new technology, far more flexible and challenging than thermometers. Understanding computers and integrating them into science is far harder than the tortured development of modern thermometry. Computation not only expands science's paradigms and supports new discoveries (particularly with AI), but it also \emph{does\/} new science --- almost all modern laboratory instrumentation, including thermometers, is heavily computerized. Relying on computer models has become routine. The dependency of modern science on computers is far more tangled and complex than the now-resolved dependence on reliable temperature measurement; computers affect how scientists in all disciplines \emph{think}.

Pushing the boundaries of science, then, now involves pushing the boundaries of computer science. The synergy runs deep: for instance, while particle physics relies on powerful supercomputers, quantum physics itself is developing more powerful quantum computing.

``Computational science'' has come to mean a particular style of science based on developing and using explicit computational models, but, really, \emph{all\/} of science is now computational in this sense.

Computational science is not just restricted to specialized fields like computational chemistry, genomics, big data~\ldots\ in all fields of science, computation is used at every step, from calculations of course, through note taking, sound and image processing, literature searches, analysis and statistics, correspondence with co-authors and editors, through to typesetting,  distributing and archiving the final publications. All areas of science are being profoundly computerized. Furthermore, developments in computer science themselves drive science such that earlier science is even becoming obsolete as the computer technology moves on \cite{form}.

\subsection{Code quality concerns}\label{related-work}
Publishing high quality computer code has been strongly advocated since the earliest times, such as the \emph{Communications of the ACM\/} in its first issue in its first volume published in 1958, where it outlined its algorithm publication policy. The new policy was illustrated with a square root algorithm \cite{acm-algorithms}. However, publishing code in the computer science literature is distinct from publishing high quality general science that depends on code, which is the particular concern of the present paper. Of course, as a special case, Computer Science too can also benefit from improving ways to reliably use code in general science.

In almost all published science, the code it relies on is taken for granted, just as in routine chemistry the quality of the glassware is not at issue. While chemists are trained in reliable methodologies, so taking quality glassware for granted is reasonable. Code is a newer innovation, and quality code is a current research programme in its own right, and has resulted in calls for a Grand Challenge research effort \cite{hoare-correctness}. Inevitably, because of these reasons taken collectively, much published science depends on unreliable code that is not explicitly discussed, was not peer-reviewed, and is not open to scrutiny, reproduction, or reuse by the scientific community. The situation with code is analogous to chemists using contaminated glassware with no awareness how its effects might be controlled or might affect results.

Is it a problem? A study of \digitBlock{863878} Python-coded Jupyter notebooks \cite{jupyter-study} found a 76\% failure rate for code to complete execution successfully. Trisovic \emph{et al.}\ \cite{trisovic} performed a  study of \digitBlock{9000} research codes written in the language R on Dataverse, an open-source repository maintained by Harvard University's Institute for Quantitative Social Sciences. They found a comparable result that 74\% of the code files analyzed failed. These results are consistent with this paper's findings, as summarized in tables \ref{table-overview} \& \ref{table-summary}.

The authors of \cite{jupyter-study} make technical recommendations to improve reproducibility, such as ``Abstract code into functions, classes, and modules and test them.'' It will not be obvious for most practicing scientists how to do this, so, more generally, the authors of \cite{trisovic} recommend establishing Working Groups to support reproducible research --- and in fact Dataverse (the source of the \cite{trisovic} data) has already done so. These ideas may be compared to the present paper's proposal of Software Engineering Boards (SEBs), as discussed in detail in section \ref{summary}, supported by more specific suggestions in the \supplement.

These concerns about code are part of the reproducibility crisis for science generally \cite{machine-learning-reproducibility,no-raw-data,data-access,reproducible-manifesto,reproducibility-crisis}. Concern has led to new journals, \emph{Journal of Open Source Software\/} (JOSS) \cite{joss}, \emph{ReScience~C\/} \cite{rescience}, and  others, to explore and encourage the explicit replication of previously published research.

To start to address code quality issues the Transparency and Openness Promotion Committee met in 2014, and has since been promoting Transparency and Openness Promotion, TOP, starting with journal publication policies \cite{TOP}. TOP covers citation standards, replication standards, and code standards amongst others. TOP recommends levels of compliance to their recommendations, where level 0 does not meet the standard, and levels 1 to 3 are increasingly stringent. The TOP levels for code are shown in table \ref{TOPtable}. The TOP standards continue to develop, and are now maintained on a wiki at \url{osf.io/9f6gx/wiki/Guidelines} \cite{TOP-wiki}. TOP can be compared with the Findable, Accessible, Interoperable and Reusable (FAIR) initiative, which places greater emphasis on data rather than code; FAIR itself is critiqued in section \ref{critique-fair}.

As the present paper argues, developing quality code is widely under-appreciated, which leads to a vicious cycle of lack of acknowledgement, invisibility, and being unable to recruit adequately competent coders (see section \ref{section-vicious-cycles}). The term \emph{research software engineer\/} was coined in 2012 to help address this problem, and to stimulate thinking about researchers' career paths. The Society of Research Software Engineering (\url{https://society-rse.org}) has been established to further promote research software engineer interests.

There is no shortage of computational tools available to help address the problems. However, it should be noted that such tools are not a panacea, as the study \cite{jupyter-study} cited above makes clear. This suggests that human support for improving coding and reproduction quality, perhaps in the form of Working Groups or Boards, as this paper suggests, will be critical.

In addition to unintentional problems with code quality and reproducibility, actual scientific misconduct occurs when the outcome is intentional. While pure plagiarism, which is a form of misconduct, generally does not affect the quality of reported science, when data or code is fraudulently manipulated to have deceptive properties, the results are likely to be destructive. The Wakefield MMR fraud claiming to link vaccines and autism published in \emph{The Lancet\/} took 12 years before it was retracted; this misconduct has been extraordinarily destructive \cite{MMR-fraud}.

A recent meta-analysis of surveys of scientific misconduct estimates that nearly 2\% of scientists have at least once fabricated, falsified, or modified results, and over a third have undertaken other questionable research practices \cite{misconduct-issues}. The meta-analysis qualifies the figures carefully, but these are alarming rates regardless of the qualifications; indeed, the authors suggest that as misconduct is a sensitive issue, the rates are likely to be under-estimates.\footnote{The present paper's author's own survey of scientists publishing in the \emph{Journal of Machine Learning\/} had comparable results \cite{machine-learning-reproducibility}.}

While technical solutions like using AI may help, it is notable that many misconduct issues can be detected and constructively managed prior to publication using exactly the same methods as will improve research reproducibility, as discussed throughout the present paper.

\subsection{Computable papers}\label{computable-papers}
``Electronic lab notebooks'' (ELNs) \cite{elns}, emphasize computer tools that specifically support laboratory notebook authoring and editing. In contrast, so-called ``computable papers'' aim to support the scientific paper authoring workflow: the emphasis is that papers should produce faithful results from code embedded in (or easily accessible from) the text of the paper.

If, for example, HTML is being used for a computational paper, the paper's text could include Javascript code like

{\tt <script>\par \hskip 2em document.write(responses.total)\par </script>}

This illustrative code would insert the result of running it into the paper, perhaps like ``We collected data and obtained 754 responses,'' where the 754 is the value of the variable \texttt{responses.total} when the paper was formatted. The point is that the number 754 (or whatever) is computed automatically from the data, so any changes or improvements to the data or the methodology analyzing it will translate into the published number being updated and inserted into the paper at the appropriate point.

\def\factorial#1{\digitBlock{\ffactorial{#1}{1}}}
\newcount\t
\def\ffactorial#1#2{%
\ifnum#1=0 #2\else
\ffactorial{\the\numexpr#1-1\relax}{\the\numexpr#1*#2\relax}%
\fi}

Systems like \LaTeX\ (which was used for the current paper) can combine advanced typesetting with computable paper calculations. Here, as a simple example, we calculate that
$10!$ is equal to \factorial{10} just by writing the \LaTeX\ code \verb|\factorial{10}| \emph{in\/} this paper to calculate arbitrary factorials and group the result into conventional blocks of three digits by inserting small spaces. Note that although \texttt{factorial} is not a standard \LaTeX\ function, it was readily defined by code also \emph{in\/} the present paper, so it can be easily modified by the author for other purposes as required.

More practically for helping author complex scientific papers, a separate program can generate a file of \LaTeX\ definitions for data, tables, and cross-references, etc, as needed for a paper. Here is a small example, where each generated fact (taken from the present paper's analysis) has been highlighted in \textbf{bold}:

\vskip 1ex\newdimen\boxwidth
\boxwidth = \columnwidth
\advance\boxwidth by - 1em
\noindent\framebox[\columnwidth][r]{\parbox{\boxwidth}{\noindent\sf For the present paper, as a concrete example of this approach, the pilot survey analyzed \textbf{\the\dataN} papers with \textbf{\the\countAuthors} authors. The paper \textbf{\cite{ICmodel}}, discussed below in section \textbf{\ref{section-pandemic-modeling}}, relied on code composed of \textbf{\covidsimfiles} files, with over \textbf{\covidsimkLOC\ thousand} lines of code (how this compares with files in the survey is summarized in table \textbf{\ref{supplementaryTableRepoSummary}} in the \supplement).}}\vskip 1ex

In the ``old days'' such numbers (\the\dataN; \the\countAuthors; etc) quoted above would have been manually worked out, eyeballed, typed up, and then (hopefully) double-checked by the authors. Instead, in this paper, those numbers (and many more) were computed automatically. They were inserted into the text of this paper automatically, and are auditable back to their sources. Furthermore, they will update automatically --- with no further work from the author --- when the data or calculations change.

The idea is easy to implement. Continuing using \LaTeX\ as the illustrative word processing system, code could generate text like

\begin{center}\texttt{\textbackslash newcommand\{\textbackslash numberOfAuthors\}\{\the\countAuthors\}}\end{center}

where the \the\countAuthors\ is the calculated number, save that line of text to a file, which is then import it into the paper where it is typeset in the normal way. Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the actual value is at the time. Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11}, \ldots, \digitBlock{1000},  \digitBlock{1001}, etc, as done (more realistically) in this paper's footnote~\numbernamesexamplefootnote.

The key concept in computable papers is that as the authors of a paper collect or revise data or calculations or otherwise update it, the results in the paper also update automatically, and all the statistics, graphs, and analysis reported in the paper remain correct \emph{with no further work from the authors}. Indeed, if an author corrects a mistake, the correction will apply automatically to all future revisions.

However, there is no structure to using general-purpose systems like HTML, \LaTeX, or Rmarkdown, which means authors may make unnoticed mistakes. Many authors therefore prefer a more structured approach imposed by tools designed for the purpose, where their structure helps impose uniformity and helps prevent and check for errors.

A taut review of such systems is Perkel \cite{perkel-review}, whereas \cite{maneage} discusses in-depth the design trade-offs of one powerful approach, Maneage; the paper \cite{maneage} includes (in an appendix) a substantial and useful literature review. Here four representative tools (WEB, \emph{Mathematica}, Jupyter, and knitr) serve to sample the variety of approaches that are available:

@@

\hskip 1em WEB documents an entire program, but there are variants such as relit \cite{relit} that allow arbitrary parts of programs to be documented, and hence are useful for normal scientific papers that need to explain algorithms, but do not need to show or explain the entire code required for computer execution.

\hskip 1em In contrast to the other tools reviewed here, literate programming is intended to produce high quality publications \emph{about\/} code, rather than publications just \emph{using\/} code, inserting output, such as statistics or graphs, generated by running it.

\item
\emph{Mathematica\/} is one of the earliest fully interactive notebook tools for computational papers. Notebooks were developed by Theodore Gray in 1988 \cite{notebook-patent}. Notebooks consist of a collection of ``cells,'' where cells can be labelled as text or as a section heading, but if a cell is labelled as code, it can be run and it will normally generate a new cell following it as its output. The new cell can be numbers, tables, mathematics, a plot, an image, or even more text --- the arbitrary output of running code, in fact. Moreover, a notebook itself is a \emph{Mathematica\/} expression, so it --- the paper itself --- can be analyzed or manipulated by code in any way. The entire notebook structure and contents can be checked for consistency and correctness in any way the author chooses.
